---
title: "maldives_manuscript"
output: html_document
date: "2026-04-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Load Packages

```{r}
library(tidyverse) 
library(ggplot2) 
library(vegan) 
library(dplyr)
library(scales) 
library(grid) 
library(phyloseq) 
library(grafify) 
library(microbiome) 
library(dunn.test) 
library(stats) 
library(ggordiplots) 
library(DescTools) 
library(ANCOMBC) 
library(ggrepel)    
library(compositions)
library(writexl)
library(patchwork)
library(stringr)
library(openxlsx)
library(tidytext) 
library(ggraph)
library(tidygraph)
library(phyloseq)
library(dplyr)
library(tidyr)
library(igraph)
library(Hmisc)   
library(knitr)
library(ape)
library(phangorn)   
library(dplyr)
library(tibble)
library(rstatix)
library(ggvenn)
library(eulerr)
library(ggpubr)
library(ggvenn)
library(eulerr)
library(ggforce)
```


#Working directory and Load Data
```{r}
dirS <- "/Users/lara_kalisch/Desktop/revised_maldives_manuscript"
setwd(dirS)
getwd()

## THEME SETTING####
theme_set(theme_bw())
### IMPORTING AND PREPARING DATA####
ASVsS <- read.table("ASVs_counts-no-contam.tsv",
                     header = TRUE,
                     row.names = 1,
                     sep = "\t",
                     check.names = FALSE)

### Import taxonomy table
taxfileS <- read.table("ASVs_taxonomy-no-contam.tsv",
                      header = TRUE,
                      row.names = 1,
                      sep = "\t",
                      check.names = FALSE)

### Import metadata
mapfileS <- read.csv("meta_maldives.csv",
                     header = TRUE,
                     row.names = 1,
                     check.names = FALSE)


# sanity check
table(mapfileS$sample_type, useNA = "ifany")
```

#Phyloseq
```{r}
# Ensure correct object classes
taxfileS <- as.matrix(taxfileS)
mapfileS <- as.data.frame(mapfileS)
ASVsS <- as.matrix(ASVsS)

# Create phyloseq object
phy_mw <- phyloseq(
  otu_table(ASVsS, taxa_are_rows = TRUE),
  tax_table(taxfileS),
  sample_data(mapfileS)
)

# Remove non-bacterial contaminants and singletons
phy_mw <- phy_mw %>%
  subset_taxa(
    Kingdom == "Bacteria" &
      Family != "Mitochondria" &
      Order != "Chloroplast"
  ) %>%
  prune_taxa(taxa_sums(.) > 1, .)

# Standardize sample_type metadata
sample_data(phy_mw)$sample_type <- sample_data(phy_mw)$sample_type |>
  as.character() |>
  trimws() |>
  tolower() |>
  factor(levels = c("colon", "cloaca"))

# Save cleaned full object
saveRDS(phy_mw, file = "phy_mw_clean_full.rds")

# Extract turtle ID from sample names and subset to EI < 49
ei_num <- as.numeric(sub("^EI([0-9]+).*", "\\1", sample_names(phy_mw)))

phy_mw <- phy_mw %>%
  prune_samples(ei_num < 49, .) %>%
  prune_samples(sample_sums(.) > 0, .) %>%
  prune_taxa(taxa_sums(.) > 0, .)

# Save filtered object
saveRDS(phy_mw, file = "phy_mw_pre49.rds")

phy_mw

```

#Rarifaction
```{r}

otu_mw <- otu_table(phy_mw)
otu_mw <- as.data.frame(t(as(otu_mw, "matrix")))  

vegan::rarecurve(
  otu_mw,
  step  = 10000,
  label = TRUE
)

# Check sequencing depth
summary(sample_sums(phy_mw))
table(sample_sums(phy_mw) < 50000)
low_depth_ids <- sample_names(phy_mw)[sample_sums(phy_mw) < 50000]
table(sample_data(phy_mw)[low_depth_ids, "sample_type"])

# Rarefy at the minimum sequencing depth to retain all samples,
# as higher thresholds would disproportionately exclude cloacal samples

set.seed(42)
rar_mw <- rarefy_even_depth(
  phy_mw,
  sample.size = 45815,
  rngseed     = 42,
  replace     = FALSE,
  verbose     = TRUE
)

# Remove taxa (ASVs) with zero counts after rarefaction
rar_mw <- prune_taxa(taxa_sums(rar_mw) > 0, rar_mw)

# Quick checks
rar_mw
ntaxa(rar_mw)
summary(sample_sums(rar_mw))
levels(sample_data(rar_mw)$sample_type)

# Save rarefied object
saveRDS(rar_mw, "rar_mw_pre49.rds")

### Basic info
ntaxa(phy_mw)
nsamples(phy_mw)
summary(sample_sums(phy_mw))
 
```
#Basic Stats

```{r}
track <- read.delim("read-count-tracking.tsv", row.names = 1, check.names = FALSE)  
raw_reads <- sum(track$dada2_input, na.rm = TRUE)
raw_reads
nonchim_reads <- sum(track$nonchim, na.rm = TRUE)
nonchim_reads
perc_retained <- (nonchim_reads / raw_reads) * 100
perc_retained
round(perc_retained, 1)

```

#Alpha Diversity 

```{r}

# 1. Calculate alpha-diversity metrics from rarefied data


alpha_mw <- estimate_richness(
  rar_mw,
  measures = c(
    "Observed",
    "Shannon",
    "Simpson"
  )
) %>%
  rownames_to_column("sample_ID")


# 2. Add sample metadata

meta_mw_df <- as(
  sample_data(rar_mw),
  "data.frame"
) %>%
  rownames_to_column("sample_ID")


alpha_mw_df <- alpha_mw %>%
  left_join(
    meta_mw_df,
    by = "sample_ID"
  ) %>%
  filter(
    sample_type %in% c(
      "colon",
      "cloaca"
    )
  ) %>%
  mutate(
    SampleType = factor(
      as.character(sample_type),
      levels = c(
        "colon",
        "cloaca"
      )
    ),
    turtle_ID = factor(
      as.character(turtle_ID)
    )
  )

# 3. Confirm paired structure and retain complete pairs

pair_check <- alpha_mw_df %>%
  count(
    turtle_ID,
    SampleType
  ) %>%
  pivot_wider(
    names_from = SampleType,
    values_from = n,
    values_fill = 0
  )


complete_pair_ids <- pair_check %>%
  filter(
    colon == 1,
    cloaca == 1
  ) %>%
  pull(turtle_ID)


alpha_mw_df <- alpha_mw_df %>%
  filter(
    turtle_ID %in% complete_pair_ids
  )


cat(
  "Number of complete turtle pairs:",
  length(complete_pair_ids),
  "\n"
)

print(
  table(alpha_mw_df$SampleType)
)


stopifnot(
  length(complete_pair_ids) > 0,
  all(
    c("colon", "cloaca") %in%
      levels(alpha_mw_df$SampleType)
  )
)


# 4. Create wide table for paired statistical testing

alpha_wide <- alpha_mw_df %>%
  select(
    turtle_ID,
    SampleType,
    Observed,
    Shannon,
    Simpson
  ) %>%
  pivot_wider(
    names_from = SampleType,
    values_from = c(
      Observed,
      Shannon,
      Simpson
    )
  ) %>%
  drop_na()


cat(
  "Complete pairs used in statistical tests:",
  nrow(alpha_wide),
  "\n"
)


# 5. Create long table for plotting and summaries

metric_levels <- c(
  "Observed",
  "Shannon",
  "Simpson"
)


alpha_long <- alpha_mw_df %>%
  select(
    sample_ID,
    turtle_ID,
    SampleType,
    all_of(metric_levels)
  ) %>%
  pivot_longer(
    cols = all_of(metric_levels),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    Metric = factor(
      Metric,
      levels = metric_levels
    )
  )


# 6. Calculate medians, quartiles and IQRs

alpha_summary <- alpha_long %>%
  group_by(
    Metric,
    SampleType
  ) %>%
  summarise(
    n = sum(!is.na(Value)),
    
    Median = median(
      Value,
      na.rm = TRUE
    ),
    
    Q1 = stats::quantile(
      Value,
      probs = 0.25,
      na.rm = TRUE,
      names = FALSE
    ),
    
    Q3 = stats::quantile(
      Value,
      probs = 0.75,
      na.rm = TRUE,
      names = FALSE
    ),
    
    IQR = stats::IQR(
      Value,
      na.rm = TRUE
    ),
    
    .groups = "drop"
  )


print(alpha_summary)



# 7. Function for paired Wilcoxon tests and effect sizes
#   +1 = all non-zero paired differences favour cloaca
#   -1 = all non-zero paired differences favour colon


paired_alpha_summary <- function(
    colon_values,
    cloaca_values,
    metric_name
) {
  
  keep <- complete.cases(
    colon_values,
    cloaca_values
  )
  
  colon_values <- colon_values[keep]
  cloaca_values <- cloaca_values[keep]
  
  paired_difference <-
    cloaca_values - colon_values
  
  nonzero_difference <-
    paired_difference[
      paired_difference != 0
    ]
  
  if (length(nonzero_difference) == 0) {
    
    rank_biserial <- 0
    
  } else {
    
    signed_ranks <- rank(
      abs(nonzero_difference),
      ties.method = "average"
    )
    
    positive_rank_sum <- sum(
      signed_ranks[
        nonzero_difference > 0
      ]
    )
    
    negative_rank_sum <- sum(
      signed_ranks[
        nonzero_difference < 0
      ]
    )
    
    rank_biserial <- (
      positive_rank_sum -
        negative_rank_sum
    ) / (
      positive_rank_sum +
        negative_rank_sum
    )
  }
  
  wilcox_output <- wilcox.test(
    x = cloaca_values,
    y = colon_values,
    paired = TRUE,
    exact = FALSE,
    alternative = "two.sided"
  )
  
  tibble(
    Metric = metric_name,
    n_pairs = length(colon_values),
    
    Colon_median = median(
      colon_values,
      na.rm = TRUE
    ),
    Colon_Q1 = quantile(
      colon_values,
      0.25,
      na.rm = TRUE,
      names = FALSE
    ),
    Colon_Q3 = quantile(
      colon_values,
      0.75,
      na.rm = TRUE,
      names = FALSE
    ),
    Colon_IQR = IQR(
      colon_values,
      na.rm = TRUE
    ),
    
    Cloaca_median = median(
      cloaca_values,
      na.rm = TRUE
    ),
    Cloaca_Q1 = quantile(
      cloaca_values,
      0.25,
      na.rm = TRUE,
      names = FALSE
    ),
    Cloaca_Q3 = quantile(
      cloaca_values,
      0.75,
      na.rm = TRUE,
      names = FALSE
    ),
    Cloaca_IQR = IQR(
      cloaca_values,
      na.rm = TRUE
    ),
    
    Median_paired_difference =
      median(
        paired_difference,
        na.rm = TRUE
      ),
    
    Wilcoxon_V =
      unname(
        wilcox_output$statistic
      ),
    
    p_value =
      wilcox_output$p.value,
    
    Rank_biserial =
      rank_biserial
  )
}


# 8. Run paired tests for all metrics

alpha_stats <- bind_rows(
  
  paired_alpha_summary(
    colon_values =
      alpha_wide$Observed_colon,
    cloaca_values =
      alpha_wide$Observed_cloaca,
    metric_name =
      "Observed"
  ),
  
  paired_alpha_summary(
    colon_values =
      alpha_wide$Shannon_colon,
    cloaca_values =
      alpha_wide$Shannon_cloaca,
    metric_name =
      "Shannon"
  ),
  
  paired_alpha_summary(
    colon_values =
      alpha_wide$Simpson_colon,
    cloaca_values =
      alpha_wide$Simpson_cloaca,
    metric_name =
      "Simpson"
  )
) %>%
  mutate(
    BH_adjusted_p = p.adjust(
      p_value,
      method = "BH"
    ),
    
    Significance = case_when(
      BH_adjusted_p < 0.0001 ~ "****",
      BH_adjusted_p < 0.001  ~ "***",
      BH_adjusted_p < 0.01   ~ "**",
      BH_adjusted_p < 0.05   ~ "*",
      TRUE                   ~ "ns"
    ),
    
    Effect_magnitude = case_when(
      abs(Rank_biserial) < 0.10 ~
        "Negligible",
      
      abs(Rank_biserial) < 0.30 ~
        "Small",
      
      abs(Rank_biserial) < 0.50 ~
        "Moderate",
      
      TRUE ~
        "Large"
    ),
    
    Effect_direction = case_when(
      Rank_biserial > 0 ~
        "Higher in cloaca",
      
      Rank_biserial < 0 ~
        "Higher in colon",
      
      TRUE ~
        "No directional effect"
    )
  )


print(alpha_stats)


# 9. Save paired differences for each turtle

alpha_paired_differences <- alpha_wide %>%
  transmute(
    turtle_ID,
    
    Observed_colon,
    Observed_cloaca,
    Observed_difference =
      Observed_cloaca -
      Observed_colon,
    
    Shannon_colon,
    Shannon_cloaca,
    Shannon_difference =
      Shannon_cloaca -
      Shannon_colon,
    
    Simpson_colon,
    Simpson_cloaca,
    Simpson_difference =
      Simpson_cloaca -
      Simpson_colon
  )


# 10. Prepare BH-adjusted figure annotations

annotation_df <- alpha_long %>%
  group_by(Metric) %>%
  summarise(
    value_min = min(
      Value,
      na.rm = TRUE
    ),
    value_max = max(
      Value,
      na.rm = TRUE
    ),
    value_range =
      value_max - value_min,
    .groups = "drop"
  ) %>%
  mutate(
    value_range = if_else(
      value_range == 0,
      abs(value_max) * 0.10 + 0.01,
      value_range
    ),
    
    y_position =
      value_max +
      0.10 * value_range
  ) %>%
  left_join(
    alpha_stats %>%
      transmute(
        Metric = factor(
          Metric,
          levels = metric_levels
        ),
        Significance,
        BH_adjusted_p
      ),
    by = "Metric"
  ) %>%
  mutate(
    annotation_label = case_when(
      BH_adjusted_p < 0.001 ~
        paste0(
          Significance,
          "\nBH p < 0.001"
        ),
      
      TRUE ~
        paste0(
          Significance,
          "\nBH p = ",
          formatC(
            BH_adjusted_p,
            format = "f",
            digits = 3
          )
        )
    )
  )


print(annotation_df)


# 11. Plot paired alpha diversity
# Lines show paired samples from the same turtle.
# Figure annotations use BH-adjusted p-values.


p_alpha <- ggplot(
  alpha_long,
  aes(
    x = SampleType,
    y = Value
  )
) +
  
  geom_line(
    aes(
      group = turtle_ID
    ),
    linewidth = 0.45,
    alpha = 0.30,
    colour = "grey45"
  ) +
  
  geom_boxplot(
    aes(
      fill = SampleType
    ),
    outlier.shape = NA,
    alpha = 0.80,
    colour = "black",
    linewidth = 0.7,
    width = 0.55
  ) +
  
  geom_point(
    aes(
      fill = SampleType
    ),
    shape = 21,
    size = 2,
    alpha = 0.80,
    colour = "black",
    stroke = 0.4,
    position = position_jitter(
      width = 0.035,
      height = 0,
      seed = 42
    )
  ) +
  
  geom_text(
    data = annotation_df,
    aes(
      x = 1.5,
      y = y_position,
      label = annotation_label
    ),
    inherit.aes = FALSE,
    size = 4.2,
    fontface = "bold",
    lineheight = 1.1
  ) +
  
  facet_wrap(
    ~Metric,
    scales = "free_y",
    nrow = 1
  ) +
  
  scale_fill_manual(
    values = c(
      "colon" = "#009E73",
      "cloaca" = "#7B4EA3"
    )
  ) +
  
  scale_x_discrete(
    labels = c(
      "colon" = "Colon",
      "cloaca" = "Cloaca"
    )
  ) +
  
  scale_y_continuous(
    expand = expansion(
      mult = c(
        0.05,
        0.20
      )
    )
  ) +
  
  theme_classic(
    base_size = 15
  ) +
  
  theme(
    legend.position = "none",
    
    panel.grid =
      element_blank(),
    
    axis.text =
      element_text(
        colour = "black"
      ),
    
    axis.title =
      element_text(
        colour = "black"
      ),
    
    axis.text.x =
      element_text(
        size = 14
      ),
    
    strip.text =
      element_text(
        face = "bold",
        colour = "black",
        size = 16,
        margin = margin(
          b = 8
        )
      ),
    
    strip.background =
      element_blank(),
    
    plot.title =
      element_text(
        margin = margin(
          b = 10
        )
      ),
    
    plot.subtitle =
      element_text(
        margin = margin(
          b = 14
        )
      ),
    
    panel.spacing =
      grid::unit(
        1.5,
        "lines"
      )
  ) +
  
  labs(
    title =
      "Alpha diversity of colon and cloacal bacterial communities",
    
    subtitle =
      "Lines connect paired samples from the same turtle",
    
    x = NULL,
    y = NULL
  )


print(p_alpha)

# 12. per-sample outpu

alpha_per_sample <- alpha_mw_df %>%
  select(
    sample_ID,
    turtle_ID,
    SampleType,
    Observed,
    Shannon,
    Simpson
  ) %>%
  arrange(
    turtle_ID,
    SampleType
  )


# 13. Export all requested statistics

write_xlsx(
  list(
    Median_IQR_summary =
      alpha_summary,
    
    Paired_tests_effect_sizes =
      alpha_stats,
    
    Paired_differences =
      alpha_paired_differences,
    
    Per_sample_alpha_values =
      alpha_per_sample
  ),
  path =
    "Supplementary_alpha_diversity_statistics.xlsx"
)

# 14. Save updated figure


ggsave(
  filename =
    "MW_alpha_diversity_paired_no_Chao1.png",
  plot =
    p_alpha,
  width = 13,
  height = 5.5,
  dpi = 600,
  bg = "white"
)


ggsave(
  filename =
    "MW_alpha_diversity_paired_no_Chao1.svg",
  plot =
    p_alpha,
  width = 13,
  height = 5.5,
  device = "svg",
  bg = "white"
)

write.csv(
  alpha_summary,
  "MW_alpha_median_IQR_summary.csv",
  row.names = FALSE
)


write.csv(
  alpha_stats,
  "MW_alpha_paired_tests_effect_sizes.csv",
  row.names = FALSE
)


write.csv(
  alpha_per_sample,
  "MW_alpha_per_sample_values.csv",
  row.names = FALSE
)


# 16. Final console summary

cat(
  "\nAlpha-diversity analysis complete.\n",
  "Complete pairs:",
  nrow(alpha_wide),
  "\n\n"
)

print(
  alpha_stats %>%
    select(
      Metric,
      n_pairs,
      Colon_median,
      Colon_Q1,
      Colon_Q3,
      Cloaca_median,
      Cloaca_Q1,
      Cloaca_Q3,
      Median_paired_difference,
      Wilcoxon_V,
      p_value,
      BH_adjusted_p,
      Rank_biserial,
      Effect_magnitude,
      Effect_direction
    )
)


```


#Beta Diversity
##Aitchision
```{r}
##  Aitchison distance (CLR-transformed ASV table)

# Subset to colon and cloaca samples only
phy_beta_mw <- phy_mw %>%
  subset_samples(sample_type %in% c("colon", "cloaca")) %>%
  prune_samples(sample_sums(.) > 0, .) %>%
  prune_taxa(taxa_sums(.) > 0, .)

# Basic checks
stopifnot(sum(is.na(sample_data(phy_beta_mw)$sample_type)) == 0)
stopifnot(nsamples(phy_beta_mw) == 96)

# CLR transform
phy_beta_mw_clr <- transform(phy_beta_mw, "clr")

# Extract CLR matrix (samples × ASVs)
otu_mat <- as(otu_table(phy_beta_mw_clr), "matrix")
if (taxa_are_rows(phy_beta_mw_clr)) {
  otu_mat <- t(otu_mat)
}

stopifnot(identical(rownames(otu_mat), sample_names(phy_beta_mw_clr)))

# Aitchison distance
dist_aitch_mw <- dist(otu_mat, method = "euclidean")

# Metadata
meta_mw <- data.frame(sample_data(phy_beta_mw_clr))

# PERMANOVA (paired design blocked by turtle_ID)
set.seed(42)
adonis_aitch_mw <- adonis2(
  dist_aitch_mw ~ sample_type,
  data = meta_mw,
  permutations = 999,
  strata = meta_mw$turtle_ID
)

adonis_aitch_mw

# Homogeneity of dispersion
set.seed(42)
disp_aitch_mw <- betadisper(dist_aitch_mw, meta_mw$sample_type)
permutest(disp_aitch_mw, permutations = 999)
anova(disp_aitch_mw)

## PCA for visualization
ord <- prcomp(otu_mat, center = TRUE, scale. = FALSE)
eig <- ord$sdev^2 / sum(ord$sdev^2) * 100

# Clean sample type labels
clean_sample_type <- function(x) {
  x %>%
    as.character() %>%
    stringr::str_replace_all("\u00A0", " ") %>%
    stringr::str_replace_all("\\s+", " ") %>%
    trimws() %>%
    tolower()
}

# Sample scores
df_plot <- data.frame(
  sample_ID = rownames(otu_mat),
  turtle_ID = meta_mw$turtle_ID,
  PC1 = ord$x[, 1],
  PC2 = ord$x[, 2],
  sample_type = meta_mw$sample_type
) %>%
  mutate(
    SampleType = factor(
      clean_sample_type(sample_type),
      levels = c("colon", "cloaca")
    )
  )

# Sanity checks
stopifnot(nrow(df_plot) == 96)
stopifnot(sum(is.na(df_plot$SampleType)) == 0)

# PCA loadings
loadings <- as.data.frame(ord$rotation) %>%
  tibble::rownames_to_column("ASV")

tax <- as.data.frame(tax_table(phy_beta_mw_clr)) %>%
  tibble::rownames_to_column("ASV")

# Select top genus/family drivers
top_taxa <- loadings %>%
  left_join(tax, by = "ASV") %>%
  mutate(
    Genus  = ifelse(tolower(Genus) %in% c("unclassified", "unknown", ""), NA, Genus),
    Family = ifelse(tolower(Family) %in% c("unclassified", "unknown", ""), NA, Family),
    label  = ifelse(!is.na(Genus), Genus, Family),
    label  = stringr::str_replace_all(label, "_", " ")
  ) %>%
  filter(!is.na(label)) %>%
  group_by(label) %>%
  slice_max(order_by = abs(PC1), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(desc(abs(PC1))) %>%
  slice(1:6)

# Scale arrows for plotting
arrow_mult <- quantile(abs(c(df_plot$PC1, df_plot$PC2)), 0.9) * 8

top_taxa <- top_taxa %>%
  mutate(
    PC1_plot = PC1 * arrow_mult,
    PC2_plot = PC2 * arrow_mult
  )

# Plot
p_MW <- ggplot(df_plot, aes(PC1, PC2)) +
  stat_ellipse(
    aes(colour = SampleType),
    type = "t",
    level = 0.95,
    linewidth = 0.8
  ) +
  geom_point(
    aes(colour = SampleType),
    size = 3,
    alpha = 0.9
  ) +
  geom_segment(
    data = top_taxa,
    aes(x = 0, y = 0, xend = PC1_plot, yend = PC2_plot),
    arrow = arrow(length = unit(0.25, "cm")),
    linewidth = 0.8,
    colour = "black",
    inherit.aes = FALSE
  ) +
  ggrepel::geom_text_repel(
    data = top_taxa,
    aes(x = PC1_plot, y = PC2_plot, label = label),
    size = 3.8,
    fontface = "italic",
    box.padding = 0.5,
    segment.color = NA,
    bg.color = "white",
    bg.r = 0.15,
    max.overlaps = 50,
    inherit.aes = FALSE
  ) +
  scale_colour_manual(
    values = c(
      "colon"  = "#009E73",
      "cloaca" = "#7B4EA3"
    )
  ) +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank()) +
  labs(
    title = "Aitchison PCA of Wild Maldives Sea Turtles",
    x = paste0("PC1 (", round(eig[1], 1), "%)"),
    y = paste0("PC2 (", round(eig[2], 1), "%)"),
    colour = "Sample type"
  )

p_MW

# Save figure
ggsave(
  "MW_Aitchison_PCA_colon_vs_cloaca_genus_loadings.png",
  p_MW,
  width = 10,
  height = 8,
  dpi = 600,
  bg = "white"
)

ggsave(
  "MW_Aitchison_PCA_colon_vs_cloaca_genus_loadings.svg",
  p_MW,
  width = 10,
  height = 8,
  device = "svg"
)

# Save loading tables
loading_table <- loadings %>%
  left_join(tax, by = "ASV") %>%
  mutate(
    Genus = ifelse(is.na(Genus) | Genus == "", "Unclassified", Genus),
    abs_PC1 = abs(PC1),
    abs_PC2 = abs(PC2)
  ) %>%
  arrange(desc(abs_PC1)) %>%
  select(ASV, Genus, Family, PC1, PC2, abs_PC1, abs_PC2)

write.csv(
  loading_table,
  "MW_Aitchison_PCA_Genus_Loadings.csv",
  row.names = FALSE
)

write.csv(
  loading_table %>% slice(1:20),
  "MW_Aitchison_PCA_Top20_Genus_Loadings.csv",
  row.names = FALSE
)
```


#Taxonomy
##Phyla

```{r}

phy_phylum <- phy_mw %>%
  prune_taxa(taxa_sums(.) > 0, .) %>%
  tax_glom(taxrank = "Phylum", NArm = FALSE)

# Convert to relative abundance
phy_phylum_rel <- transform_sample_counts(phy_phylum, function(x) x / sum(x))

# Melt to long format and clean labels
df_phylum <- psmelt(phy_phylum_rel) %>%
  mutate(
    Phylum = as.character(Phylum),
    Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", Phylum),
    sample_type = tolower(trimws(as.character(sample_type)))
  )

## Color palette
base_pastel <- c(
  "#9AD0EC", "#FBC4AB", "#B5EAD7", "#FFAFCC", "#CDB4DB",
  "#E4C1F9", "#FFF1A8", "#CED4DA", "#BEE7E8", "#FFD6A5",
  "#CAFFBF", "#FDFFB6", "#FFC6FF", "#BDB2FF", "#EAE4E9",
  "#F1FAEE", "#E9C46A", "#F4A261", "#84A59D", "#B8F2E6"
)

## Top phyla for plotting
topN <- 10

top_phyla <- df_phylum %>%
  group_by(Phylum) %>%
  summarise(mean_abund = mean(Abundance), .groups = "drop") %>%
  arrange(desc(mean_abund)) %>%
  slice_head(n = topN) %>%
  pull(Phylum)

df_phylum <- df_phylum %>%
  mutate(
    Phylum_plot = ifelse(Phylum %in% top_phyla, Phylum, "Other")
  )

phyla_levels <- sort(unique(df_phylum$Phylum_plot))
phyla_levels <- phyla_levels[phyla_levels != "Other"]

phyla_colors <- setNames(
  base_pastel[seq_along(phyla_levels)],
  phyla_levels
)

phyla_colors <- c(phyla_colors, Other = "grey85")
phyla_colors["Pseudomonadota"] <- "#8DA0CB"

## Mean composition by sample type
df_phylum_mean <- df_phylum %>%
  group_by(Sample, sample_type, Phylum_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Sample, sample_type) %>%
  mutate(Abundance = Abundance / sum(Abundance)) %>%
  ungroup() %>%
  group_by(sample_type, Phylum_plot) %>%
  summarise(Abundance = mean(Abundance), .groups = "drop")

p_phylum_mean <- ggplot(
  df_phylum_mean,
  aes(x = sample_type, y = Abundance, fill = Phylum_plot)
) +
  geom_col(color = "black", linewidth = 0.2) +
  scale_fill_manual(values = phyla_colors) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(limits = c("colon", "cloaca")) +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank()) +
  labs(
    title = "Phylum-level gut microbiome composition (mean)",
    subtitle = paste0("Top ", topN, " phyla; remaining collapsed as Other"),
    x = NULL,
    y = "Mean relative abundance",
    fill = "Phylum"
  )

ggsave(
  "MW_Phylum_mean_colon_vs_cloaca.png",
  p_phylum_mean,
  width = 10,
  height = 6,
  dpi = 400
)

ggsave(
  "MW_Phylum_mean_colon_vs_cloaca.svg",
  p_phylum_mean,
  width = 10,
  height = 6
)

## Per-sample composition
df_phylum_ind <- df_phylum %>%
  group_by(Sample, sample_type, turtle_ID, Phylum_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  mutate(
    turtle_num = as.numeric(gsub("\\D+", "", turtle_ID)),
    Sample_label = paste0(turtle_ID, "_", sample_type)
  )

sample_order <- df_phylum_ind %>%
  distinct(Sample_label, turtle_num, sample_type) %>%
  arrange(sample_type, turtle_num) %>%
  pull(Sample_label)

df_phylum_ind$Sample_label <- factor(df_phylum_ind$Sample_label, levels = sample_order)

p_phylum_all <- ggplot(
  df_phylum_ind,
  aes(x = Sample_label, y = Abundance, fill = Phylum_plot)
) +
  geom_col(color = "black", linewidth = 0.15) +
  scale_fill_manual(values = phyla_colors) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~sample_type, scales = "free_x", nrow = 2) +
  theme_bw(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5),
    axis.ticks.x = element_blank()
  ) +
  labs(
    title = "Phylum-level composition per sample",
    subtitle = "Samples ordered by turtle ID; colon and cloaca shown separately",
    x = "Turtle ID",
    y = "Relative abundance",
    fill = "Phylum"
  )

ggsave(
  "MW_Phylum_individual_all.png",
  p_phylum_all,
  width = 14,
  height = 8,
  dpi = 400,
  bg = "white"
)

ggsave(
  "MW_Phylum_individual_all.svg",
  p_phylum_all,
  width = 14,
  height = 8,
  device = "svg"
)


## Export phylum summary tables
top_phyla_tables <- df_phylum %>%
  group_by(sample_type, Phylum) %>%
  summarise(
    mean_abundance = mean(Abundance, na.rm = TRUE),
    sd_abundance = sd(Abundance, na.rm = TRUE),
    n_samples = sum(!is.na(Abundance)),
    se_abundance = sd_abundance / sqrt(n_samples),
    .groups = "drop"
  ) %>%
  group_by(sample_type) %>%
  arrange(desc(mean_abundance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 20) %>%
  ungroup()

write.csv(
  top_phyla_tables,
  "MW_Top20_Phyla_per_sample_type.csv",
  row.names = FALSE
)


## Paired Wilcoxon tests by phylum
df_phylum_clean <- df_phylum %>%
  group_by(turtle_ID, sample_type, Phylum) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

df_phylum_wide <- df_phylum_clean %>%
  tidyr::pivot_wider(
    names_from = sample_type,
    values_from = Abundance
  ) %>%
  filter(!is.na(colon) & !is.na(cloaca))

phylum_wilcox_results <- lapply(unique(df_phylum_wide$Phylum), function(p) {
  sub <- df_phylum_wide[df_phylum_wide$Phylum == p, ]

  test <- wilcox.test(
    sub$colon,
    sub$cloaca,
    paired = TRUE
  )

  data.frame(
    Phylum = p,
    p_value = test$p.value
  )
}) %>%
  dplyr::bind_rows()

phylum_wilcox_results$p_adj <- p.adjust(phylum_wilcox_results$p_value, method = "BH")

phylum_means <- df_phylum_clean %>%
  group_by(sample_type, Phylum) %>%
  summarise(mean_abundance = mean(Abundance), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = sample_type, values_from = mean_abundance)

phylum_stats <- phylum_wilcox_results %>%
  left_join(phylum_means, by = "Phylum") %>%
  mutate(
    higher_in = ifelse(colon > cloaca, "colon", "cloaca")
  ) %>%
  arrange(p_adj)

phylum_stats

writexl::write_xlsx(
  phylum_stats,
  "MW_Wild_Phyla_Paired_Wilcoxon.xlsx"
)

#Merge tables
# Reshape abundance summary table to wide format
phylum_summary_wide <- top_phyla_tables %>%
  select(sample_type, Phylum, mean_abundance, sd_abundance, se_abundance) %>%
  pivot_wider(
    names_from = sample_type,
    values_from = c(mean_abundance, sd_abundance, se_abundance)
  )

# Merge summary abundances with paired Wilcoxon results
phylum_supp_table <- phylum_summary_wide %>%
  left_join(
    phylum_stats %>%
      select(Phylum, p_value, p_adj, higher_in),
    by = "Phylum"
  ) %>%
  arrange(p_adj)

# View merged table
phylum_supp_table

# Save merged supplementary table
write.csv(
  phylum_supp_table,
  "MW_SuppTable_Phyla_Abundance_and_Wilcoxon.csv",
  row.names = FALSE
)

writexl::write_xlsx(
  phylum_supp_table,
  "MW_SuppTable_Phyla_Abundance_and_Wilcoxon.xlsx"
)
```


##Families
```{r}

# Agglomerate at family level
phy_family <- phy_mw %>%
  prune_taxa(taxa_sums(.) > 0, .) %>%
  tax_glom(taxrank = "Family", NArm = FALSE)

# Convert to relative abundance
phy_family_rel <- transform_sample_counts(phy_family, function(x) x / sum(x))

# Melt and clean labels
df_family <- psmelt(phy_family_rel) %>%
  mutate(
    Family = as.character(Family),
    Family = ifelse(is.na(Family) | Family == "", "Unclassified", Family),
    sample_type = tolower(trimws(as.character(sample_type)))
  )

## Top families for plotting
topN <- 20

top_fams <- df_family %>%
  group_by(Family) %>%
  summarise(mean_abund = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_abund)) %>%
  slice_head(n = topN) %>%
  pull(Family)

df_family <- df_family %>%
  mutate(
    Family_plot = ifelse(Family %in% top_fams, Family, "Other")
  )

## Family colors
family_levels <- sort(unique(df_family$Family_plot))
family_levels <- family_levels[family_levels != "Other"]

family_colors <- setNames(
  base_pastel[seq_along(family_levels)],
  family_levels
)

family_colors <- c(family_colors, Other = "grey85")
family_colors["Marinifilaceae"] <- "#8DA0CB"

## Mean composition by sample type
df_family_mean <- df_family %>%
  group_by(Sample, sample_type, Family_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Sample, sample_type) %>%
  mutate(Abundance = Abundance / sum(Abundance)) %>%
  ungroup() %>%
  group_by(sample_type, Family_plot) %>%
  summarise(Abundance = mean(Abundance), .groups = "drop")
df_family_mean <- df_family_mean %>%
  mutate(
    Family_plot = factor(
      Family_plot,
      levels = c("Other", setdiff(unique(Family_plot), "Other"))
    )
  )

p_family_mean <- ggplot(
  df_family_mean,
  aes(x = sample_type, y = Abundance, fill = Family_plot)
) +
  geom_col(color = "black", linewidth = 0.2) +
  scale_fill_manual(values = family_colors) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(limits = c("colon", "cloaca")) +
  theme_bw(base_size = 14) +
  theme(panel.grid = element_blank()) +
  labs(
    title = "Family-level gut microbiome composition (mean)",
    subtitle = paste0("Top ", topN, " families shown; remaining collapsed as Other"),
    x = NULL,
    y = "Mean relative abundance",
    fill = "Family"
  )

ggsave(
  "MW_Family_mean_colon_vs_cloaca.png",
  p_family_mean,
  width = 12,
  height = 7,
  dpi = 400,
  bg = "white"
)

ggsave(
  "MW_Family_mean_colon_vs_cloaca.svg",
  p_family_mean,
  width = 12,
  height = 7,
  device = "svg"
)

## Per-sample composition
df_family_ind <- df_family %>%
  group_by(Sample, sample_type, turtle_ID, Family_plot) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  mutate(
    turtle_num = as.numeric(gsub("\\D+", "", turtle_ID)),
    Sample_label = paste0(turtle_ID, "_", sample_type)
  )

sample_order <- df_family_ind %>%
  distinct(Sample_label, turtle_num, sample_type) %>%
  arrange(turtle_num, sample_type) %>%
  pull(Sample_label)

df_family_ind$Sample_label <- factor(df_family_ind$Sample_label, levels = sample_order)

p_family_all <- ggplot(
  df_family_ind,
  aes(x = Sample_label, y = Abundance, fill = Family_plot)
) +
  geom_col(color = "black", linewidth = 0.15) +
  scale_fill_manual(values = family_colors) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~sample_type, scales = "free_x", nrow = 2) +
  theme_bw(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5),
    axis.ticks.x = element_blank()
  ) +
  labs(
    title = "Family-level composition per sample",
    subtitle = "Samples ordered by turtle ID; colon and cloaca shown separately",
    x = "Turtle ID",
    y = "Relative abundance",
    fill = "Family"
  )

ggsave(
  "MW_Family_individual_all.png",
  p_family_all,
  width = 14,
  height = 8,
  dpi = 400,
  bg = "white"
)

ggsave(
  "MW_Family_individual_all.svg",
  p_family_all,
  width = 14,
  height = 8,
  device = "svg"
)

## Full family summary table
family_summary_full <- df_family %>%
  group_by(sample_type, Family) %>%
  summarise(
    mean_abundance = mean(Abundance, na.rm = TRUE),
    sd_abundance = sd(Abundance, na.rm = TRUE),
    n_samples = n_distinct(Sample),
    se_abundance = sd_abundance / sqrt(n_samples),
    .groups = "drop"
  )

## Paired Wilcoxon tests by family
df_family_clean <- df_family %>%
  group_by(turtle_ID, sample_type, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

df_family_wide <- df_family_clean %>%
  tidyr::pivot_wider(
    names_from = sample_type,
    values_from = Abundance
  ) %>%
  filter(!is.na(colon) & !is.na(cloaca))

family_stats <- lapply(unique(df_family_wide$Family), function(fam) {
  sub <- df_family_wide[df_family_wide$Family == fam, ]

  test <- wilcox.test(
    sub$colon,
    sub$cloaca,
    paired = TRUE,
    exact = FALSE
  )

  data.frame(
    Family = fam,
    p_value = test$p.value
  )
}) %>%
  dplyr::bind_rows()

family_stats$p_adj <- p.adjust(family_stats$p_value, method = "BH")

family_means <- df_family_clean %>%
  group_by(sample_type, Family) %>%
  summarise(mean_abundance = mean(Abundance), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from = sample_type,
    values_from = mean_abundance
  )

family_stats <- family_stats %>%
  left_join(family_means, by = "Family") %>%
  mutate(
    higher_in = ifelse(colon > cloaca, "colon", "cloaca"),
    diff = colon - cloaca,
    abs_diff = abs(diff)
  ) %>%
  arrange(p_adj)

## Merge full abundance summary with Wilcoxon results
family_summary_wide <- family_summary_full %>%
  select(sample_type, Family, mean_abundance, sd_abundance, se_abundance) %>%
  tidyr::pivot_wider(
    names_from = sample_type,
    values_from = c(mean_abundance, sd_abundance, se_abundance)
  )

family_supp_table <- family_summary_wide %>%
  left_join(
    family_stats %>%
      select(Family, p_value, p_adj, higher_in),
    by = "Family"
  ) %>%
  arrange(p_adj)

# Save final supplementary table
write.csv(
  family_supp_table,
  "MW_SuppTable_Families_Abundance_and_Wilcoxon.csv",
  row.names = FALSE
)

writexl::write_xlsx(
  family_supp_table,
  "MW_SuppTable_Families_Abundance_and_Wilcoxon.xlsx"
)

## Top 20 significant family differences
family_top20 <- family_stats %>%
  filter(p_adj < 0.05) %>%
  arrange(desc(abs_diff)) %>%
  slice_head(n = 20)

family_top20
```

##Genus
```{r}
# -----------------------------
# 1. Prepare genus-level data
# -----------------------------
topN <- 30

phy_genus_rel <- phy_mw %>%
  subset_samples(sample_type %in% c("colon", "cloaca")) %>%
  prune_samples(sample_sums(.) > 0, .) %>%
  prune_taxa(taxa_sums(.) > 0, .) %>%
  tax_glom(taxrank = "Genus", NArm = FALSE) %>%
  transform_sample_counts(function(x) x / sum(x))

df_genus <- psmelt(phy_genus_rel) %>%
  mutate(
    Genus = as.character(Genus),
    Genus = if_else(is.na(Genus) | Genus == "", "Unclassified", Genus),
    sample_type = factor(sample_type, levels = c("colon", "cloaca"))
  )


# 2. Summary stats + Wilcoxon 

# agglomaerat per turtle
df_genus_clean <- df_genus %>%
  group_by(turtle_ID, sample_type, Genus) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

# Summary stats (mean ± SE)
genus_summary <- df_genus_clean %>%
  group_by(sample_type, Genus) %>%
  summarise(
    mean_abundance = mean(Abundance),
    sd_abundance   = sd(Abundance),
    n_samples      = n(),
    se_abundance   = sd_abundance / sqrt(n_samples),
    .groups = "drop"
  ) %>%
  mutate(
    mean_pct = mean_abundance * 100,
    se_pct   = se_abundance * 100
  )

# Pivot summary to wide (colon vs cloaca side-by-side)
genus_summary_wide <- genus_summary %>%
  select(sample_type, Genus, mean_pct, se_pct) %>%
  pivot_wider(
    names_from = sample_type,
    values_from = c(mean_pct, se_pct)
  )

# -----------------------------
# Paired Wilcoxon
# -----------------------------
df_genus_wide <- df_genus_clean %>%
  pivot_wider(
    names_from  = sample_type,
    values_from = Abundance
  ) %>%
  drop_na(colon, cloaca)

genus_wilcox <- df_genus_wide %>%
  group_by(Genus) %>%
  group_modify(~{
    test <- wilcox.test(.x$colon, .x$cloaca, paired = TRUE)
    tibble(p_value = test$p.value)
  }) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"))

# Mean values for direction + effect size
genus_means <- df_genus_wide %>%
  group_by(Genus) %>%
  summarise(
    colon = mean(colon),
    cloaca = mean(cloaca),
    mean_diff = abs(colon - cloaca),
    .groups = "drop"
  ) %>%
  mutate(
    higher_in = if_else(colon > cloaca, "colon", "cloaca")
  )


# final table
genus_final <- genus_summary_wide %>%
  left_join(genus_wilcox, by = "Genus") %>%
  left_join(genus_means, by = "Genus") %>%
  arrange(desc(mean_diff))

# Top 50 + All genera

top50_genera <- genus_final %>%
  slice_head(n = 50)


# Export
writexl::write_xlsx(
  list(
    Top50 = top50_genera,
    All_genera = genus_final
  ),
  "MW_Genus_Final_COLON_vs_CLOACA.xlsx"
)

```



#ANCOM
## Paired ANCOM ASV level
```{r}

# Paired ANCOM-BC2 analysis:
# Reference level: cloaca
# lfc_sample_typecolon:  positive = colon-enriched  negative = cloaca-enriched


# 1. subset


phy_asv <- phy_mw %>%
  subset_samples(
    species == "EI" &
      grepl("^wild", status, ignore.case = TRUE) &
      sample_type %in% c("colon", "cloaca") &
      !is.na(sample_type) &
      !is.na(turtle_ID)
  )

phy_asv <- prune_samples(
  sample_sums(phy_asv) > 0,
  phy_asv
)

phy_asv <- prune_taxa(
  taxa_sums(phy_asv) > 0,
  phy_asv
)



# 2. set factors

sample_data(phy_asv)$sample_type <- factor(
  as.character(sample_data(phy_asv)$sample_type),
  levels = c("cloaca", "colon")
)

sample_data(phy_asv)$turtle_ID <- factor(
  as.character(sample_data(phy_asv)$turtle_ID)
)


# 3. Confirm each turtle has colon and cloaca sample


pair_table <- table(
  sample_data(phy_asv)$turtle_ID,
  sample_data(phy_asv)$sample_type
)

print(pair_table)

complete_pairs <- rownames(pair_table)[
  pair_table[, "cloaca"] == 1 &
    pair_table[, "colon"] == 1
]

cat(
  "\nNumber of complete turtle pairs:",
  length(complete_pairs),
  "\n"
)

cat("Samples retained:", nsamples(phy_asv), "\n")
cat("ASVs before ANCOM-BC2 filtering:", ntaxa(phy_asv), "\n")
print(table(sample_data(phy_asv)$sample_type))

# 4. paired ANCOM-BC2
# sample_type = fixed effect
# turtle_ID   = random intercept

set.seed(123)

ancom_asv <- ancombc2(
  data = phy_asv,
  assay_name = "otu_table",
  tax_level = NULL,
  fix_formula = "sample_type",
  rand_formula = "(1 | turtle_ID)",
  group = "sample_type",
  p_adj_method = "BH",
  prv_cut = 0.10,
  lib_cut = 1000,
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  verbose = TRUE
)

# Save object
saveRDS(
  ancom_asv,
  "ANCOMBC2_paired_colon_vs_cloaca_model.rds"
)


# 5. Extract modeled ANCOM-BC2 results


                # Verify columns
required_columns <- c(
  
  "taxon",
  
  "lfc_sample_typecolon",
  
  "se_sample_typecolon",
  
  "p_sample_typecolon",
  
  "q_sample_typecolon",
  
  "passed_ss_sample_typecolon",
  
  "diff_sample_typecolon",
  
  "diff_robust_sample_typecolon"
  
)

missing_columns <- setdiff(
  
  required_columns,
  
  colnames(ancom_asv$res)
  
)

if (length(missing_columns) > 0) {
  
  stop(
    
    "Missing expected result columns: ",
    
    paste(missing_columns, collapse = ", ")
    
  )
  
}

           # extract results

required_columns <- c(
  "taxon",
  "lfc_sample_typecolon",
  "se_sample_typecolon",
  "p_sample_typecolon",
  "q_sample_typecolon",
  "passed_ss_sample_typecolon",
  "diff_sample_typecolon",
  "diff_robust_sample_typecolon"
)

missing_columns <- setdiff(
  required_columns,
  colnames(ancom_asv$res)
)

if (length(missing_columns) > 0) {
  stop(
    "Missing expected result columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

res_asv <- ancom_asv$res %>%
  as.data.frame() %>%
  transmute(
    ASV = as.character(taxon),
    logFC = lfc_sample_typecolon,
    se = se_sample_typecolon,
    p_val = p_sample_typecolon,
    q_val = q_sample_typecolon,
    passed_ss = passed_ss_sample_typecolon,
    diff_abn = diff_sample_typecolon,
    diff_robust = diff_robust_sample_typecolon
  )



# 6. Attach taxonomy 

tax_df <- as.data.frame(
  tax_table(phy_asv),
  stringsAsFactors = FALSE
) %>%
  rownames_to_column("ASV")

res_asv <- res_asv %>%
  left_join(tax_df, by = "ASV") %>%
  mutate(
    across(
      any_of(c("Phylum", "Class", "Order", "Family", "Genus")),
      ~na_if(as.character(.x), "Unclassified")
    ),
    Label = case_when(
      !is.na(Genus) & Genus != "" ~
        paste0(Genus, " (", ASV, ")"),
      
      !is.na(Family) & Family != "" ~
        paste0("Unclassified ", Family, " (", ASV, ")"),
      
      !is.na(Order) & Order != "" ~
        paste0("Unclassified ", Order, " (", ASV, ")"),
      
      !is.na(Class) & Class != "" ~
        paste0("Unclassified ", Class, " (", ASV, ")"),
      
      !is.na(Phylum) & Phylum != "" ~
        paste0("Unclassified ", Phylum, " (", ASV, ")"),
      
      TRUE ~ ASV
    ),
    
    # Direction depends only on coefficient sign.
    direction = case_when(
      logFC > 0 ~ "Colon-enriched",
      logFC < 0 ~ "Cloaca-enriched",
      TRUE ~ NA_character_
    )
  )



# 7. Create result set containing significant after BH and robust sensitivity (ANCOM SS FILTER)


# BH-significant modeled ASVs
sig_asv <- res_asv %>%
  filter(
    diff_abn %in% TRUE,
    !is.na(logFC)
  ) %>%
  arrange(q_val)

# Significant ASVs that also pass pseudo-count sensitivity
sig_asv_robust <- res_asv %>%
  filter(
    diff_robust %in% TRUE,
    !is.na(logFC)
  ) %>%
  arrange(q_val)



# 8. Summarize results

cat("\nASVs modeled:", nrow(res_asv), "\n")
cat("BH-significant modeled ASVs:", nrow(sig_asv), "\n")
cat(
  "Pseudo-count-robust significant ASVs:",
  nrow(sig_asv_robust),
  "\n"
)

cat("\nBH-significant ASVs by direction:\n")
print(sig_asv %>% count(direction))

cat("\nRobust significant ASVs by direction:\n")
print(sig_asv_robust %>% count(direction))


# 9. Extract structural-zero ASVs (many of them might be bioloigcal important as they are almost absent in one of the sample types)


zero_asv <- ancom_asv$zero_ind %>%
  as.data.frame() %>%
  mutate(
    ASV = as.character(taxon),
    
    direction = case_when(
      `structural_zero (sample_type = cloaca)` %in% TRUE &
        !(`structural_zero (sample_type = colon)` %in% TRUE) ~
        "Colon-associated",
      
      `structural_zero (sample_type = colon)` %in% TRUE &
        !(`structural_zero (sample_type = cloaca)` %in% TRUE) ~
        "Cloaca-associated",
      
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(direction)) %>%
  select(-taxon) %>%
  left_join(tax_df, by = "ASV")

cat("\nStructural-zero ASVs by direction:\n")
print(zero_asv %>% count(direction))


# 10. Add prevalence and abundance to structural zeros to determine the most important ones 
# Subset to structural-zero ASVs before converting to long format to reduce memory use.


otu_zero <- as(
  otu_table(phy_asv),
  "matrix"
)

if (!taxa_are_rows(phy_asv)) {
  otu_zero <- t(otu_zero)
}

zero_ids <- intersect(
  zero_asv$ASV,
  rownames(otu_zero)
)

otu_zero <- otu_zero[zero_ids, , drop = FALSE]

meta_zero <- data.frame(
  sample_data(phy_asv),
  check.names = FALSE
) %>%
  rownames_to_column("sample_name")

zero_counts_long <- as.data.frame(otu_zero) %>%
  rownames_to_column("ASV") %>%
  pivot_longer(
    cols = -ASV,
    names_to = "sample_name",
    values_to = "count"
  ) %>%
  left_join(
    meta_zero %>%
      select(
        sample_name,
        turtle_ID,
        sample_type
      ),
    by = "sample_name"
  )

zero_summary <- zero_counts_long %>%
  group_by(ASV, sample_type) %>%
  summarise(
    n_samples = n(),
    n_detected = sum(count > 0),
    prevalence = mean(count > 0),
    mean_count = mean(count),
    median_count = median(count),
    total_count = sum(count),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = sample_type,
    values_from = c(
      n_samples,
      n_detected,
      prevalence,
      mean_count,
      median_count,
      total_count
    ),
    names_sep = "_"
  )

zero_asv_full <- zero_asv %>%
  left_join(zero_summary, by = "ASV") %>%
  mutate(
    prevalence_present_group = case_when(
      direction == "Colon-associated" ~ prevalence_colon,
      direction == "Cloaca-associated" ~ prevalence_cloaca,
      TRUE ~ NA_real_
    ),
    
    mean_count_present_group = case_when(
      direction == "Colon-associated" ~ mean_count_colon,
      direction == "Cloaca-associated" ~ mean_count_cloaca,
      TRUE ~ NA_real_
    ),
    
    total_count_present_group = case_when(
      direction == "Colon-associated" ~ total_count_colon,
      direction == "Cloaca-associated" ~ total_count_cloaca,
      TRUE ~ NA_real_
    )
  )



# 11. Rank structural-zero ASVs as they do not have p values


top_structural_zero <- zero_asv_full %>%
  arrange(
    desc(prevalence_present_group),
    desc(mean_count_present_group)
  )

top_structural_zero_by_direction <- zero_asv_full %>%
  group_by(direction) %>%
  arrange(
    desc(prevalence_present_group),
    desc(mean_count_present_group),
    .by_group = TRUE
  ) %>%
  slice_head(n = 10) %>%
  ungroup()

print(
  top_structural_zero %>%
    select(
      ASV,
      Genus,
      direction,
      prevalence_cloaca,
      prevalence_colon,
      mean_count_cloaca,
      mean_count_colon
    ) %>%
    head(20)
)


# 12. Export complete supplementary table with ancom and structural zero results


write_xlsx(
  list(
    All_modelled_ASVs = res_asv,
    Significant_modelled_ASVs = sig_asv,
    Robust_significant_ASVs = sig_asv_robust,
    All_structural_zero_ASVs = zero_asv_full,
    Top_structural_zeros = top_structural_zero,
    Top_10_zeros_by_direction =
      top_structural_zero_by_direction
  ),
  "ANCOMBC2_paired_colon_vs_cloaca_complete.xlsx"
)


############################################
# 13. Select ASVs for Figure 3C
#
# 4 modeled ASVs per direction = 8 bars
# 2 structural-zero ASVs per direction = 4 diamonds
# Total = 12 ASVs


# 13A. Select ANCOM model ASVs:  Rank by absolute estimated log fold change.


modelled_for_figure <- sig_asv_robust %>%
  filter(
    !is.na(logFC),
    !is.na(direction)
  ) %>%
  group_by(direction) %>%
  slice_max(
    order_by = abs(logFC),
    n = 4,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  mutate(
    result_type = "Robust estimated logFC",
    plot_direction = direction,
    plot_value = logFC,
    Plot_label = Label
  )



# 13B. Select structural-zero ASVs
#Rank first by prevalence, then by mean abundance

zero_for_figure <- zero_asv_full %>%
  filter(!is.na(direction)) %>%
  group_by(direction) %>%
  arrange(
    desc(prevalence_present_group),
    desc(mean_count_present_group),
    .by_group = TRUE
  ) %>%
  slice_head(n = 2) %>%
  ungroup() %>%
  mutate(
    plot_direction = recode(
      direction,
      "Colon-associated" = "Colon-enriched",
      "Cloaca-associated" = "Cloaca-enriched"
    ),
    
    Plot_label = case_when(
      !is.na(Genus) & Genus != "" ~
        paste0(Genus, " (", ASV, "; SZ)"),
      
      !is.na(Family) & Family != "" ~
        paste0(
          "Unclassified ",
          Family,
          " (",
          ASV,
          "; SZ)"
        ),
      
      TRUE ~ paste0(ASV, " (SZ)")
    ),
    
    result_type = "Structural zero"
  )



# 14. Define structural-zero marker positions
# Structural-zero markers are placed just outside the modeled logFC range. Their position indicates direction, not numerical effect size.


max_abs_logFC <- max(
  abs(modelled_for_figure$logFC),
  na.rm = TRUE
)

structural_zero_position <- max_abs_logFC * 1.15

zero_for_figure <- zero_for_figure %>%
  mutate(
    plot_value = case_when(
      plot_direction == "Colon-enriched" ~
        structural_zero_position,
      
      plot_direction == "Cloaca-enriched" ~
        -structural_zero_position,
      
      TRUE ~ NA_real_
    )
  )


# 15. Create shared row order


plot_order <- bind_rows(
  modelled_for_figure %>%
    select(
      Plot_label,
      plot_value
    ),
  
  zero_for_figure %>%
    select(
      Plot_label,
      plot_value
    )
) %>%
  arrange(plot_value)

label_levels <- plot_order$Plot_label

modelled_for_figure <- modelled_for_figure %>%
  mutate(
    Plot_label = factor(
      Plot_label,
      levels = label_levels
    )
  )

zero_for_figure <- zero_for_figure %>%
  mutate(
    Plot_label = factor(
      Plot_label,
      levels = label_levels
    )
  )



# 16. Plot combined ANCOM-BC2 results
# Bars:
#   modeled ASVs with estimated log fold changes
# Diamonds:
#   structural-zero ASVs; position shows direction only


p_ancom_asv <- ggplot() +
  
  geom_col(
    data = modelled_for_figure,
    aes(
      x = plot_value,
      y = Plot_label,
      fill = plot_direction
    ),
    width = 0.65
  ) +
  
  geom_point(
    data = zero_for_figure,
    aes(
      x = plot_value,
      y = Plot_label,
      fill = plot_direction
    ),
    shape = 23,
    size = 4,
    stroke = 0.8,
    color = "black"
  ) +
  
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey40",
    linewidth = 0.7
  ) +
  
  scale_fill_manual(
    values = c(
      "Colon-enriched" = "#4E8B5C",
      "Cloaca-enriched" = "#7B1FA2"
    ),
    na.translate = FALSE
  ) +
  
  scale_x_continuous(
    limits = c(
      -structural_zero_position * 1.08,
      structural_zero_position * 1.08
    ),
    breaks = pretty(
      c(-max_abs_logFC, max_abs_logFC),
      n = 5
    )
  ) +
  
  labs(
    title =
      "Differentially abundant ASVs between colon and cloaca",
    
    subtitle = expression(
      paste(
        "Paired ANCOM-BC2 analysis of wild ",
        italic("E. imbricata")
      )
    ),
    
    x = "Log fold change (colon relative to cloaca)",
    y = NULL,
    fill = NULL
  ) +
  
  theme_classic(base_size = 16) +
  
  theme(
    axis.text.y = element_text(size = 9),
    axis.title.x = element_text(size = 13),
    legend.position = "top"
  )

print(p_ancom_asv)


# 17. Save Figure 3C draft


ggsave(
  filename =
    "ANCOMBC2_paired_colon_vs_cloaca_8modeled_4SZ.png",
  plot = p_ancom_asv,
  width = 10,
  height = 7,
  dpi = 600,
  bg = "white"
)

ggsave(
  filename =
    "ANCOMBC2_paired_colon_vs_cloaca_8modeled_4SZ.svg",
  plot = p_ancom_asv,
  width = 10,
  height = 7,
  bg = "white"
)


# Verify plotted taxa are also passed SS filter /  robust sensitivity
modelled_for_figure %>%
  select(
    ASV,
    Label,
    logFC,
    q_val,
    passed_ss,
    diff_robust
  )

table(modelled_for_figure$diff_robust)


# 18. Save data displayed in Figure 3C


figure3c_data <- bind_rows(
  modelled_for_figure %>%
    transmute(
      ASV,
      Plot_label = as.character(Plot_label),
      result_type,
      direction = plot_direction,
      plotted_value = plot_value,
      logFC,
      q_val,
      prevalence_present_group = NA_real_,
      mean_count_present_group = NA_real_
    ),
  
  zero_for_figure %>%
    transmute(
      ASV,
      Plot_label = as.character(Plot_label),
      result_type,
      direction = plot_direction,
      plotted_value = plot_value,
      logFC = NA_real_,
      q_val = NA_real_,
      prevalence_present_group,
      mean_count_present_group
    )
)

write.csv(
  figure3c_data,
  "ANCOMBC2_Figure3C_displayed_ASVs.csv",
  row.names = FALSE
)


# 19. Save session


capture.output(
  sessionInfo(),
  file = "ANCOMBC2_paired_sessionInfo.txt"
)

```

# Venn Diagram 
```{r}
# Use same filtered phyloseq object as ANCOM
otu <- as(otu_table(phy_asv), "matrix")
if (!taxa_are_rows(phy_asv)) otu <- t(otu)

meta <- as.data.frame(sample_data(phy_asv))

colon_samples  <- rownames(meta)[meta$sample_type == "colon"]
cloaca_samples <- rownames(meta)[meta$sample_type == "cloaca"]

# ASVs present at least once in each sample type
colon_asvs <- rownames(otu)[
  rowSums(otu[, colon_samples, drop = FALSE] > 0) > 0
]

cloaca_asvs <- rownames(otu)[
  rowSums(otu[, cloaca_samples, drop = FALSE] > 0) > 0
]

# Counts
shared      <- intersect(colon_asvs, cloaca_asvs)
colon_only  <- setdiff(colon_asvs, cloaca_asvs)
cloaca_only <- setdiff(cloaca_asvs, colon_asvs)

n_colon_only  <- length(colon_only)
n_cloaca_only <- length(cloaca_only)
n_shared      <- length(shared)
total_asvs    <- length(union(colon_asvs, cloaca_asvs))

pct <- function(x) paste0(round(x / total_asvs * 100, 1), "%")

# Check numbers
data.frame(
  Category = c("Colon only", "Shared", "Cloaca only", "Total"),
  n_ASVs = c(n_colon_only, n_shared, n_cloaca_only, total_asvs),
  Percent = c(pct(n_colon_only), pct(n_shared), pct(n_cloaca_only), "100%")
)

# Manual Venn-style plot
p_venn <- ggplot() +
  # Cloaca circle: larger/top/purple
  geom_circle(
    aes(x0 = 0, y0 = 0.45, r = 1.25),
    fill = "#D8C7F0",
    color = "#7B1FA2",
    alpha = 0.55,
    linewidth = 1.4
  ) +
  # Colon circle: smaller/bottom/green
  geom_circle(
    aes(x0 = 0, y0 = -0.55, r = 0.95),
    fill = "#CFE8D6",
    color = "#4E8B5C",
    alpha = 0.55,
    linewidth = 1.4
  ) +
  annotate("text", x = 0, y = 1.9, label = "Cloaca",
           color = "#7B1FA2", fontface = "bold", size = 7) +
  annotate("text", x = 0, y = -1.65, label = "Colon",
           color = "#4E8B5C", fontface = "bold", size = 7) +
  annotate("text", x = 0, y = 0.95,
           label = paste0(n_cloaca_only, "\n(", pct(n_cloaca_only), ")"),
           size = 5.5) +
  annotate("text", x = 0, y = -0.05,
           label = paste0(n_shared, "\n(", pct(n_shared), ")"),
           size = 5.5, fontface = "bold") +
  annotate("text", x = 0, y = -1.0,
           label = paste0(n_colon_only, "\n(", pct(n_colon_only), ")"),
           size = 5.5) +
  labs(
    title = "ASV overlap between colon and cloacal samples",
    subtitle = paste0("Presence/absence of ASVs; total ASVs = ", total_asvs)
  ) +
  coord_fixed(xlim = c(-1.8, 1.8), ylim = c(-2.0, 2.2)) +
  theme_void(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 13)
  )

p_venn

ggsave(
  "ASV_overlap_colon_cloaca_venn.png",
  p_venn,
  width = 7,
  height = 6,
  dpi = 600,
  bg = "white"
)

ggsave(
  "ASV_overlap_colon_cloaca_venn.svg",
  p_venn,
  width = 7,
  height = 6,
  bg = "white"
)
```


# ##################################
# Reviewer Analysis 

#Reviewer 1:

#Reviewer 3:
##Comment 4:Sensitivity Analysis Alpha/Beta stats to compare samples from 2023 and 2024
```{r}
#1. ALPHA DIVERSITY 

# Add sampling year to rarefied phyloseq object
sample_data(rar_mw)$year <- ifelse(
  as.numeric(
    sub("^EI([0-9]+).*", "\\1", sample_names(rar_mw))
  ) <= 29,
  "2023",
  "2024"
)

sample_data(rar_mw)$year <- factor(
  sample_data(rar_mw)$year,
  levels = c("2023", "2024")
)

# Split rarefied object by year
rar_2023 <- subset_samples(rar_mw, year == "2023")
rar_2024 <- subset_samples(rar_mw, year == "2024")

# Remove taxa absent within each year subset
rar_2023 <- prune_taxa(taxa_sums(rar_2023) > 0, rar_2023)
rar_2024 <- prune_taxa(taxa_sums(rar_2024) > 0, rar_2024)

# Estimate alpha diversity
alpha_2023 <- estimate_richness(
  rar_2023,
  measures = c("Observed", "Shannon", "Simpson")
)

alpha_2024 <- estimate_richness(
  rar_2024,
  measures = c("Observed", "Shannon", "Simpson")
)

# Add metadata
alpha_2023$sample_type <- as.character(
  sample_data(rar_2023)$sample_type
)
alpha_2023$turtle_ID <- as.character(
  sample_data(rar_2023)$turtle_ID
)

alpha_2024$sample_type <- as.character(
  sample_data(rar_2024)$sample_type
)
alpha_2024$turtle_ID <- as.character(
  sample_data(rar_2024)$turtle_ID
)

# Metrics to test
metrics <- c("Observed", "Shannon", "Simpson")

# Paired Wilcoxon: 2023
results_2023 <- lapply(metrics, function(metric) {

  wide <- alpha_2023 %>%
    dplyr::select(
      turtle_ID,
      sample_type,
      dplyr::all_of(metric)
    ) %>%
    tidyr::pivot_wider(
      names_from = sample_type,
      values_from = dplyr::all_of(metric)
    ) %>%
    tidyr::drop_na(colon, cloaca)

  test <- wilcox.test(
    wide$colon,
    wide$cloaca,
    paired = TRUE
  )

  data.frame(
    Metric = metric,
    n_pairs = nrow(wide),
    p = test$p.value
  )
}) %>%
  bind_rows() %>%
  mutate(
    BH = p.adjust(p, method = "BH")
  )

# Paired Wilcoxon: 2024
results_2024 <- lapply(metrics, function(metric) {

  wide <- alpha_2024 %>%
    dplyr::select(
      turtle_ID,
      sample_type,
      dplyr::all_of(metric)
    ) %>%
    tidyr::pivot_wider(
      names_from = sample_type,
      values_from = dplyr::all_of(metric)
    ) %>%
    tidyr::drop_na(colon, cloaca)

  test <- wilcox.test(
    wide$colon,
    wide$cloaca,
    paired = TRUE
  )

  data.frame(
    Metric = metric,
    n_pairs = nrow(wide),
    p = test$p.value
  )
}) %>%
  bind_rows() %>%
  mutate(
    BH = p.adjust(p, method = "BH")
  )

# Combine results
alpha_year_results <- bind_rows(
  results_2023 %>% mutate(Year = "2023"),
  results_2024 %>% mutate(Year = "2024")
) %>%
  select(
    Year,
    Metric,
    n_pairs,
    p,
    BH
  )

print(alpha_year_results)






#2. BETA DIVERSITY 
phy_2023 <- subset_samples(phy_mw, year == "2023")
phy_2024 <- subset_samples(phy_mw, year == "2024")

clr_2023 <- microbiome::transform(phy_2023, "clr")
clr_2024 <- microbiome::transform(phy_2024, "clr")

dist_2023 <- vegan::vegdist(t(otu_table(clr_2023)), method = "euclidean")
dist_2024 <- vegan::vegdist(t(otu_table(clr_2024)), method = "euclidean")

#permanoca
meta23 <- data.frame(sample_data(phy_2023))
meta24 <- data.frame(sample_data(phy_2024))

set.seed(123)

adonis23 <- adonis2(
  dist_2023 ~ sample_type,
  data = meta23,
  permutations = 9999,
  strata = meta23$turtle_ID
)

adonis24 <- adonis2(
  dist_2024 ~ sample_type,
  data = meta24,
  permutations = 9999,
  strata = meta24$turtle_ID
)

adonis23
adonis24

#permdisp
disp23 <- betadisper(dist_2023, meta23$sample_type)
anova(disp23)

disp24 <- betadisper(dist_2024, meta24$sample_type)
anova(disp24)

```


##Comment 5: Sequencing summary table
```{r}

# 1. Read read-tracking table

track <- read.delim(
  "read-count-tracking-wtrawreads.tsv",
  row.names = 1,
  check.names = FALSE
) %>%
  rownames_to_column("Sample")


# Keep EI1–EI48
track_96 <- track %>%
  filter(grepl("^EI[0-9]+", Sample)) %>%
  mutate(
    turtle_number = as.numeric(
      sub("^EI([0-9]+).*", "\\1", Sample)
    ),
    Sample_type = if_else(
      grepl("CL$", Sample, ignore.case = TRUE),
      "Cloaca",
      "Colon"
    )
  ) %>%
  filter(
    !is.na(turtle_number),
    turtle_number >= 1,
    turtle_number <= 48
  )


# Confirm 96 libraries
stopifnot(nrow(track_96) == 96)


# 2. Add final reads per sample from phy_mw
#nonchim = final reads after DADA2
# Final_reads = reads remaining in the cleaned phyloseq object

final_reads <- tibble(
  Sample = sample_names(phy_mw),
  Final_reads = as.numeric(sample_sums(phy_mw))
)

per_sample_summary <- track_96 %>%
  select(
    Sample,
    Sample_type,
    raw_reads,
    dada2_input,
    filtered,
    dada_f,
    dada_r,
    merged,
    nonchim
  ) %>%
  rename(
    Raw_reads = raw_reads,
    After_primer_adapter_removal = dada2_input,
    After_quality_filtering = filtered,
    Denoised_forward = dada_f,
    Denoised_reverse = dada_r,
    Merged_reads = merged,
    Non_chimeric_reads = nonchim
  ) %>%
  left_join(final_reads, by = "Sample")


# Check that all 96 samples matched
stopifnot(
  sum(is.na(per_sample_summary$Final_reads)) == 0
)


# 3. Rebuild original unfiltered phyloseq object
#
# ASVsS, taxfileS and mapfileS must still contain the original,
# unfiltered data used to create phy_mw.

phy_original <- phyloseq(
  otu_table(as.matrix(ASVsS), taxa_are_rows = TRUE),
  tax_table(as.matrix(taxfileS)),
  sample_data(as.data.frame(mapfileS))
)

# 4. Count taxonomic categories removed

tax_original <- as.data.frame(
  tax_table(phy_original),
  stringsAsFactors = FALSE
)

# Convert taxonomy columns to plain character
tax_original[] <- lapply(
  tax_original,
  function(x) trimws(as.character(x))
)

n_initial_asvs <- ntaxa(phy_original)

n_mitochondria <- sum(
  tax_original$Family == "Mitochondria",
  na.rm = TRUE
)

n_chloroplast <- sum(
  tax_original$Order == "Chloroplast",
  na.rm = TRUE
)

n_archaea <- sum(
  tax_original$Kingdom == "Archaea",
  na.rm = TRUE
)

n_eukaryotes <- sum(
  tax_original$Kingdom %in%
    c("Eukaryota", "Eukaryotes"),
  na.rm = TRUE
)


# 5. Reproduce your actual taxonomic and singleton filtering

phy_after_taxonomy <- phy_original %>%
  subset_taxa(
    Kingdom == "Bacteria" &
      Family != "Mitochondria" &
      Order != "Chloroplast"
  )

n_after_taxonomy <- ntaxa(phy_after_taxonomy)

phy_after_singletons <- phy_after_taxonomy %>%
  prune_taxa(taxa_sums(.) > 1, .)

n_after_singletons <- ntaxa(phy_after_singletons)

n_singletons_removed <-
  n_after_taxonomy - n_after_singletons



# 6. Restrict to EI1–EI48


sample_ids <- sample_names(phy_after_singletons)

ei_number <- suppressWarnings(
  as.numeric(
    sub("^EI([0-9]+).*", "\\1", sample_ids)
  )
)

keep_samples <- grepl("^EI[0-9]+", sample_ids) &
  !is.na(ei_number) &
  ei_number >= 1 &
  ei_number <= 48

phy_recreated_96 <- prune_samples(
  keep_samples,
  phy_after_singletons
)

phy_recreated_96 <- prune_samples(
  sample_sums(phy_recreated_96) > 0,
  phy_recreated_96
)

phy_recreated_96 <- prune_taxa(
  taxa_sums(phy_recreated_96) > 0,
  phy_recreated_96
)

# 7. Check recreated object against final phy_mw


cat("Recreated samples:", nsamples(phy_recreated_96), "\n")
cat("Current phy_mw samples:", nsamples(phy_mw), "\n")

cat("Recreated ASVs:", ntaxa(phy_recreated_96), "\n")
cat("Current phy_mw ASVs:", ntaxa(phy_mw), "\n")

if (
  nsamples(phy_recreated_96) != nsamples(phy_mw) ||
  ntaxa(phy_recreated_96) != ntaxa(phy_mw)
) {
  warning(
    "The recreated object does not exactly match phy_mw. ",
    "Check before using the filtering counts."
  )
}


# 8. ASV filtering summary
# Antony confirmed decontam removed zero ASVs.

n_asvs_removed_after_subset <-
  ntaxa(phy_after_singletons) - ntaxa(phy_recreated_96)

n_asvs_removed_after_subset



asv_filtering_summary <- tibble(
  Processing_step = c(
    "Initial ASVs before taxonomic filtering",
    "Contaminant ASVs removed by decontam",
    "Mitochondrial ASVs removed",
    "Chloroplast ASVs removed",
    "Archaeal ASVs removed",
    "Eukaryotic ASVs removed",
    "Singleton ASVs removed",
    "ASVs absent after restriction to EI1–EI48",
    "Final ASVs retained in EI1–EI48 dataset"
  ),
  Number_of_ASVs = c(
    n_initial_asvs,
    0,
    n_mitochondria,
    n_chloroplast,
    n_archaea,
    n_eukaryotes,
    n_singletons_removed,
    n_asvs_removed_after_subset,
    ntaxa(phy_mw)
  )
)
  


# 9. Rarefaction summary

rarefaction_depth <- 45815

n_before_rarefaction <- nsamples(phy_mw)

n_retained_rarefaction <- sum(
  sample_sums(phy_mw) >= rarefaction_depth
)

rarefaction_summary <- tibble(
  Rarefaction_depth = rarefaction_depth,
  Samples_before_rarefaction = n_before_rarefaction,
  Samples_retained_after_rarefaction =
    n_retained_rarefaction,
  Samples_removed_by_rarefaction =
    n_before_rarefaction - n_retained_rarefaction
)


# 10. Negative-control and decontam summary


decontam_summary <- tibble(
  Item = c(
    "Decontam method",
    "Decontam threshold",
    "Negative controls included",
    "B_250918 non-chimeric reads",
    "LabEnv_250914 non-chimeric reads",
    "ASVs identified as contaminants",
    "ASVs removed as contaminants"
  ),
  Result = c(
    "Prevalence",
    "0.1",
    "2",
    "14906",
    "17222",
    "0",
    "0"
  )
)


# 11. Overall read totals

overall_read_summary <- tibble(
  Processing_step = c(
    "Raw reads",
    "After primer/adaptor removal",
    "After quality filtering",
    "Merged reads",
    "Non-chimeric reads",
    "Final reads in cleaned phyloseq object"
  ),
  Total_reads = c(
    sum(track_96$raw_reads),
    sum(track_96$dada2_input),
    sum(track_96$filtered),
    sum(track_96$merged),
    sum(track_96$nonchim),
    sum(sample_sums(phy_mw))
  )
)


# 12. Export supplementary workbook

write_xlsx(
  list(
    Per_sample_read_summary = per_sample_summary,
    Overall_read_summary = overall_read_summary,
    ASV_filtering_summary = asv_filtering_summary,
    Rarefaction_summary = rarefaction_summary,
    Decontam_summary = decontam_summary
  ),
  "Supplementary_sequencing_summary.xlsx"
)



```


## Comment 6. Bray Curtis 
```{r}

# 1. Create Bray-Curtis distance matrix

bray_dist <- phyloseq::distance(
  phy_mw,
  method = "bray"
)

# 2. Prepare metadata in matching sample order

meta_bray <- as(
  sample_data(phy_mw),
  "data.frame"
)

meta_bray <- meta_bray[
  attr(bray_dist, "Labels"),
  ,
  drop = FALSE
]

stopifnot(
  identical(
    rownames(meta_bray),
    attr(bray_dist, "Labels")
  )
)


# 3. Paired PERMANOVA
# Permutations restricted within turtle ID

set.seed(123)

adonis_bray <- vegan::adonis2(
  bray_dist ~ sample_type,
  data = meta_bray,
  permutations = 9999,
  strata = meta_bray$turtle_ID
)

print(adonis_bray)


# 4. PERMDISP

disp_bray <- vegan::betadisper(
  bray_dist,
  group = meta_bray$sample_type
)

anova_bray_disp <- anova(disp_bray)

print(anova_bray_disp)

# Optional permutation test for dispersion
set.seed(123)

permutest_bray_disp <- permutest(
  disp_bray,
  permutations = 9999
)

print(permutest_bray_disp)
```

## Comment 8: Venn diagram 

```{r}
# ASV overlap sensitivity analysis
# Prevalence filters: 5% and 10%

otu_mat <- as(
  otu_table(phy_mw),
  "matrix"
)

# Ensure ASVs are rows
if (!taxa_are_rows(phy_mw)) {
  otu_mat <- t(otu_mat)
}

meta_df <- as(
  sample_data(phy_mw),
  "data.frame"
)

colon_samples <- rownames(meta_df)[
  meta_df$sample_type == "colon"
]

cloaca_samples <- rownames(meta_df)[
  meta_df$sample_type == "cloaca"
]

stopifnot(
  length(colon_samples) == 48,
  length(cloaca_samples) == 48
)


# 2. Calculate prevalence within each sample type

prevalence_table <- tibble(
  ASV = rownames(otu_mat),

  n_colon = rowSums(
    otu_mat[, colon_samples, drop = FALSE] > 0
  ),

  n_cloaca = rowSums(
    otu_mat[, cloaca_samples, drop = FALSE] > 0
  )
) %>%
  mutate(
    prevalence_colon =
      n_colon / length(colon_samples),

    prevalence_cloaca =
      n_cloaca / length(cloaca_samples)
  )



# 3. Function to calculate filtered overlap
# ASVs are retained when prevalence reaches the thresholdin at least one sample type.


calculate_overlap <- function(
    prevalence_df,
    threshold
) {

  filtered <- prevalence_df %>%
    filter(
      prevalence_colon >= threshold |
        prevalence_cloaca >= threshold
    ) %>%
    mutate(
      detected_colon = n_colon > 0,
      detected_cloaca = n_cloaca > 0,

      overlap_group = case_when(
        detected_colon & detected_cloaca ~
          "Shared",

        detected_colon & !detected_cloaca ~
          "Colon only",

        !detected_colon & detected_cloaca ~
          "Cloaca only",

        TRUE ~
          "Neither"
      )
    )

  summary <- filtered %>%
    count(
      overlap_group,
      name = "Number_of_ASVs"
    ) %>%
    mutate(
      Prevalence_threshold =
        paste0(threshold * 100, "%"),

      Percentage =
        100 * Number_of_ASVs /
        sum(Number_of_ASVs)
    )

  list(
    filtered_asvs = filtered,
    summary = summary
  )
}


# 4. Run 5% and 10% prevalence filters

overlap_5 <- calculate_overlap(
  prevalence_table,
  threshold = 0.05
)

overlap_10 <- calculate_overlap(
  prevalence_table,
  threshold = 0.10
)


# View summaries
print(overlap_5$summary)
print(overlap_10$summary)


# 5. Combine summaries

overlap_summary <- bind_rows(
  overlap_5$summary,
  overlap_10$summary
) %>%
  arrange(
    Prevalence_threshold,
    overlap_group
  )

print(overlap_summary)


# 6. Export results

write.csv(
  overlap_summary,
  "ASV_overlap_prevalence_filtered_summary.csv",
  row.names = FALSE
)

write.csv(
  overlap_5$filtered_asvs,
  "ASV_overlap_5percent_filtered_ASVs.csv",
  row.names = FALSE
)

write.csv(
  overlap_10$filtered_asvs,
  "ASV_overlap_10percent_filtered_ASVs.csv",
  row.names = FALSE
)

# 7. Update plot using 10% prevalance filter 


n_colon_only  <- 1410
n_shared      <- 1540
n_cloaca_only <- 7454

total_asvs <- n_colon_only + n_shared + n_cloaca_only


pct <- function(x) paste0(round(x / total_asvs * 100, 1), "%")

# Bubble radius approximately proportional to total ASVs
r_colon  <- 0.75
r_cloaca <- 1.55


p_venn <- ggplot() +

  # Cloaca (purple)
  geom_circle(
    aes(x0 = 0, y0 = 0.55, r = r_cloaca),
    fill = "#D8C7F0",
    colour = "#7B1FA2",
    alpha = 0.55,
    linewidth = 1.3
  ) +

  # Colon (green)
  geom_circle(
    aes(x0 = 0, y0 = -0.75, r = r_colon),
    fill = "#CFE8D6",
    colour = "#4E8B5C",
    alpha = 0.60,
    linewidth = 1.3
  ) +

  # Titles
  annotate(
    "text",
    x = 0,
    y = 2.35,
    label = "Cloaca",
    colour = "#7B1FA2",
    fontface = "bold",
    size = 7
  ) +

  annotate(
    "text",
    x = 0,
    y = -1.85,
    label = "Colon",
    colour = "#4E8B5C",
    fontface = "bold",
    size = 7
  ) +

  # Cloaca only
  annotate(
    "text",
    x = 0,
    y = 1.25,
    label = paste0(
      n_cloaca_only,
      "\n(",
      pct(n_cloaca_only),
      ")"
    ),
    size = 5.5
  ) +

  # Shared
  annotate(
    "text",
    x = 0,
    y = 0.00,
    label = paste0(
      n_shared,
      "\n(",
      pct(n_shared),
      ")"
    ),
    fontface = "bold",
    size = 5.5
  ) +

  # Colon only
  annotate(
    "text",
    x = 0,
    y = -1.10,
    label = paste0(
      n_colon_only,
      "\n(",
      pct(n_colon_only),
      ")"
    ),
    size = 5.5
  ) +

  labs(
    title = "Overlap of prevalent ASVs between colon and cloacal samples",
    subtitle = paste0(
      "ASVs detected in at least 10% of samples (n = ",
      total_asvs,
      ")"
    )
  ) +

  coord_fixed(
    xlim = c(-2.2, 2.2),
    ylim = c(-2.1, 2.6)
  ) +

  theme_void(base_size = 15) +

  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5,
      size = 18
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 13
    )
  )

p_venn

ggsave(
  "ASV_overlap_10pct_prevalence.png",
  p_venn,
  width = 7,
  height = 6,
  dpi = 600,
  bg = "white"
)

ggsave(
  "ASV_overlap_10pct_prevalence.svg",
  p_venn,
  width = 7,
  height = 6,
  bg = "white"
)


```

